---
title: "Class 11"
author: "Kalisa Kang (PID A16741690)"
format: pdf
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39960101-40090102;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) *100
```

Now let's look at a different population. I picked the GBR (Great Britain) population.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(gbr)
gbr
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.

# Section 4: Population Scale Analysis

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)

```

Let's make a boxplot

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE) + stat_summary()
```


> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
table(expr$geno)
```

The sample size for A|A is 108, A|G is 233, and G|G is 121.

```{r}
library(dplyr)
```

```{r}
medians <- expr %>%
  group_by(geno) %>%
  summarize(median_exp = median(exp))
medians
```

The median expression level for A|A is 31.25, A|G is 25.06, and G|G is 20.07.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

Based on the box plot without any statistical analyses, the expression levels of G|G are lower than that of A|A. The SNP affects expression of ORMDL3.








